A Wait-free Queue with Polylogarithmic Step Complexity

### 2022-10-18 thesis/thesis.tex
add more related work

\paragraph{Thesis outline} The rest of the thesis is organized as follows. Chapter 2 gives an outline of related word done in the area and the motivation of our implementation. Previous lock-free queues and their common problem is presented in Section 2.1. In Section 2.2 we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects. Section 2.4 is devoted to the lower bound (COMPLETE).

In Chapter 3 we introduce a poly-logarithmic step wait-free queue. Section 3.1 we give a high-level description of our implementation. Section 3.2 contains the algorithm itself.

@@

Chapter 5 presents performance analysis for our implementation. We analyze the number of \nf{CAS} instructions out algorithm invokes and the worst-case and amortized time running of the queue.

@@

\bf{TODO} 
Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed BQ: A Lock-Free Queue with Batching. Their idea of batching allows a sequence of operations to be submitted as a batch for later execution. It supports a new notion introduced by the authors called Extended Medium Futures Linearizability.

\bf{TODO}
Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.

@@

\subsection{Restricted Queues}
Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.

\bf{TODO}
Which David?

\subsection{Universal Constructions}
Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}. A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also introduced a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}. His universal construction can be used to create any data structure, but its implementation is not practical because of using \nf{CAS} instructions on unreasonably large-sized words of memory(\bf{TODO: how large?}).

\bf{TODO: why introducing this result?}
Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea has similarities to Jayanti's construction, and they represent the value of the Fetch\&Inc using the history of successful operations.

%FIFO queues have a wide range of use in OS and applications. The current state-of-the-art queues are implemented using linked lists; that's why they have factor p in their time complexity.

%\href{https://docs.google.com/spreadsheets/d/1cL1tgXXdljkh462sMwkTVMHH_k0MBTmslIVM3xA5VS4/edit#gid=0}{Table of previous works}

%There is a connection between queues and universal constructions. We can implement a universal construction using a queue. We can store operations in the shared queue and compute operations' responses using the queue's content. Some impractical universal constructions are using big words as "...".

\subsection{Attiya Fourier Lower Bound}
\pagebreak

\pagebreak
\section{Queue Implementation} \label{DescriptQ}

@@

\subsection{Pseudocode} \label{algQ}
\begin{algorithm}
\caption{Tree Fields Description}
\begin{algorithmic}[1]
\setcounter{ALG@line}{100}
\begin{multicols}{2}

@@

\begin{algorithm}
\caption{\tt{\sl{Queue}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{200}

@@

\begin{algorithm}
\caption{Root}
\begin{algorithmic}[1]
\setcounter{ALG@line}{800}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_e(root)=E_i(root,b)$,i.e. , the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within block \nf{b} of the \nf{root}.}

@@

\begin{algorithm}
\caption{Leaf}
\begin{algorithmic}[1]
\setcounter{ALG@line}{600}

@@

\begin{algorithm}
\caption{\tt{\sl{Node}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{300}
\begin{multicols}{2}

@@

\begin{algorithm}
\caption{Node}
\begin{algorithmic}[1]
\setcounter{ALG@line}{400}

@@

>>>>

\paragraph{Thesis outline} The rest of the thesis is organized as follows. Chapter 2 gives an outline of related word done in the area and the motivation of our implementation. Previous lock-free queues and their common problem is presented in Section 2.1. In Section 2.2 we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects.

In Chapter 3 we introduce a poly-logarithmic step wait-free queue. In Section 3.1 we give a high-level description of our implementation. Section 3.2 contains the algorithm itself.

@@

Chapter 5 is devoted to performance analysis for our implementation. We analyze the number of \nf{CAS} instructions out algorithm invokes and the worst-case and amortized time running of the queue.

@@

Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue with batching. Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue. They use some properties of the size of the queue before and after a batch which is similar to apart of our work. Their queue is not wait-free, in fact if the batch sizes are 1 then the queue is like MS-queue and it supports a new notion introduced by the authors called Extended Medium Futures Linearizability.

Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses fast-path slow-path methodology. Their work is based on a circular queue using bounded memory. They design a slow phase which is wait-free bounded. If a process has no progress other processes try to help it in its \nf{CAS} loop in its slow phase. Since there can be $p$ processes helping the worst-case complexity cannot be better than linear.

@@

\subsection{Restricted Queues}

David introduced the first sublinear concurrent queue \cite{DBLP:conf/wdag/David04}. Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinte memory.
The author states that to reduce memory usage to be bounded the time per operation increases to linear.

Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}. Their algorithm time complexity is $O(\log n)$ where $n$ is the number of enqueuers. We will improve their algorithm to construct a multi-enqueuer multi-dequeuer queue.

\bf{TODO: Where should be?}
Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve logarithmic complexity is to use a tree storing the list of elements added by processes and when a process wants to do Fetch\&Inc it appends its element to the tree and return the number of elements in the tree. Their work can be descibed a queue without dequeues.

\subsection{Universal Constructions}
Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}. A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. Chandra, Jayanti and Tan  introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}. Their universal construction cannot be used to create any data structure, but a set of objects called closed objects. We mention a non-practical poly-logarithmic universal construction in page 12.

\section{Queue Implementation} \label{DescriptQ}

@@

\subsection{Pseudocode} \label{algQ}
\begin{algorithm}
\caption{Tree Fields Description}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}
\begin{multicols}{2}

@@

\begin{algorithm}
\caption{\tt{\sl{Queue}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{0}

@@

\begin{algorithm}
\caption{Root}
\begin{algorithmic}[1]
\setcounter{ALG@line}{25}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_e(root)=E_i(root,b)$,i.e. , the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within block \nf{b} of the \nf{root}.}

@@

\begin{algorithm}
\caption{Leaf}
\begin{algorithmic}[1]
\setcounter{ALG@line}{34}

@@

\begin{algorithm}
\caption{\tt{\sl{Node}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{39}
\begin{multicols}{2}

@@

\begin{algorithm}
\caption{Node}
\begin{algorithmic}[1]
\setcounter{ALG@line}{85}

@@

\paragraph{Attiya--Fouren Lower Bound}
Attiya and Fouren proved a lower bound on amortized complexity of lock-free queues from the factor of contention (number of concurrent processes). Their result says any implementation of queues, using reads,
writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$
amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}. It is an open problem to create the reduce the gap between our algorithm and Attiya--Fouren lower bound.
